Boosted Mirror Descent, Frank-Wolfe, and Herding

### 2013-09-09 icml/paper.tex
added max-ent and part of the weighted averaging argument

Consider the following loss function:
\[ L(u) = \max_{\theta \in \Theta} [\langle \theta, u \rangle - R(\theta)] \]
$R$ is some convex function that we should think of as a measure of complexity 
of $\theta$. Thus, the loss is equal to the maximum amount that $\theta$ can match 
$u$, regularized by the complexity function $R(\theta)$.

Examples:
\begin{itemize}
\item Let $R(\theta) = \langle \theta, u_0 \rangle$. Then $L(u)$ is the 
      \emph{maximum mean discrepancy} between $u$ and $u_0$ 
      relative to $\Theta$.
\item Let $R(\theta) = S^*(\theta)$ where $S^*$ is the Fenchel conjugate of 
      any strongly convex function $S$. Then $L(u) = S(u)$.
\end{itemize}

In this document we consider a family of methods for minimizing $L(u)$, 
which are based on Bach's generalization of the Frank-Wolfe algorithm and can 
be interpreted as boosted mirror descent.

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\[ L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta). \]
In the case $h(u) \equiv 0$, we can get the previous setting by 
considering $\max_{\theta \in \Theta} L(u,\theta)$. We will assume 
throughout this document that $h$ and $R$ are both convex functions.

Boosted mirror descent can be thought of as an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm:
\begin{enumerate}
\item $u_1 = 0$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
Let $\hat{u} = \frac{1}{T} \sum_{t = 1}^{T} u_t$. As long as $h$ is strongly convex, we obtain the 
bound:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/\sqrt{T}). \]
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm:
\begin{enumerate}
\item $\theta_1 = 0$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\end{enumerate}
Once again, if we let $\hat{u} = \frac{1}{T} \sum_{t=1}^T u_t$, then we obtain the following bound, which now 
only requires the convexity of $R$:
\[ \sup_{\theta \in \Theta} L(\hat{u}, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/\sqrt{T}). \]

@@

We now prove the convergence results cited in Section~\ref{sec:intro}.

@@

\begin{proposition}[Convergence of Method 1]

@@

and $u_{t+1} \in \arg\min_u h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$. 
\[ \sup_{\theta} L(\hat{u}, \theta) \leq \sup_{\theta} L(u^*, \theta) + \frac{1}{T} \sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle. \]

@@

Lemma~\ref{lem:ftl} with $l_t(u) = L(u, \theta_t)$, then we get the inequality
\[ \sum_{t=1}^T L(u_t, \theta_t) \leq \sum_{t=1}^T L(u^*, \theta_t) + \sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle. \]

@@

 &\leq \frac{1}{T} \sum_{t=1}^T L(u_t, \theta_t) \\
 &\leq \frac{1}{T} \sum_{t=1}^T L(u^*, \theta_t) + \frac{1}{T} \sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle \\
 &\leq \frac{1}{T} \sum_{t=1}^T \sup_{\theta} L(u^*, \theta) + \frac{1}{T} \sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle \\
 &= \sup_{\theta} L(u^*, \theta) + \frac{1}{T} \sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle,

@@

Suppose that $h$ is $\alpha$-strongly convex and let $R = \sup_{\theta} |\langle \theta, \theta \rangle|$. Then 

@@

\|u_t-u_{t+1}\| &\leq \frac{2}{\alpha}\|\frac{1}{t}\sum_{s \leq t}\theta_s - \frac{1}{t+1}\sum_{s \leq t+1}\theta_s\| \\
 &= \frac{2}{\alpha}\|\frac{1}{t(t+1)} \sum_{s \leq t} \theta_s - \frac{1}{t+1} \theta_{t+1}\| \\
 &\leq \frac{2}{\alpha} \left(\frac{1}{t(t+1)} \sum_{s \leq t} \|\theta_s\| + \frac{1}{t+1} \|\theta_{t+1}\|\right) \\
 &\leq \frac{4R}{\alpha (t+1)}.

@@

\sum_{t=1}^T \langle u_t - u_{t+1}, \theta_t \rangle &\leq \sum_{t=1}^T \frac{4R^2}{\alpha (t+1)} \\
 &\leq \frac{4R^2\log(T+1)}{\alpha},

@@

@@

\begin{proposition}[Convergence of Method 2]

@@

@@

\section{Application to Herding}
\label{sec:herding}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e. one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max\limits_{i=1}^d |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Method 1 and Method 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $R = \sup_{x} \|\phi(x)\|_2$; 
then using the Method 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

>>>>

Consider the following loss function:
\begin{equation}
L(u) = h(u) + \max_{\theta \in \Theta} [\langle \theta, u \rangle - R(\theta)]
\end{equation}
We can think of $h$ as a \emph{primal regularizer} and $R$ as a \emph{dual regularizer}.

Examples:
\begin{itemize}
\item Let $h(u) = 0$ and $R(\theta) = \langle \theta, u_0 \rangle$. Then $L(u)$ is the 
      \emph{maximum mean discrepancy} between $u$ and $u_0$ 
      relative to $\Theta$.
\item Let $h(u) = 0$ and $R(\theta) = S^*(\theta)$ where $S^*$ is the Fenchel conjugate of 
      any strongly convex function $S$. Then $L(u) = S(u)$.
\item Let $h(u) = \|u\|_1$, $R(u) = \langle \theta, u_0 \rangle + \frac{1}{2} \|\theta\|_2^2$. 
      Then $L(u) = \|u\|_1 + \frac{1}{2} \|u-u_0\|_2^2$. 
\end{itemize}

In this paper we will consider a family of methods for minimizing $L(u)$, 
which are based on Bach's generalization of the Frank-Wolfe algorithm and can 
be interpreted as boosted mirror descent.

First, let us generalize the setting 
to consider a \emph{two-argument} loss function
\begin{equation}
L(u,\theta) = h(u) + \langle \theta, u \rangle - R(\theta).
\end{equation}
We will assume throughout that $h$ and $R$ are both convex functions, 
and furthermore that $\arg\max_{u} h(u) + \theta^Tu$ can be efficiently 
computed for all values of $\theta$.

Mirror descent, together with boosting, yields an algorithm for finding 
``saddle points'' of $L$. It is the following algorithm (called Algorithm 1).
For notational convenience, for a sequence of weights $\alpha_1,\alpha_2,\ldots$ 
let $\hat{u}_t = \frac{\sum_{s=1}^t \alpha_su_s}{\sum_{s=1}^t \alpha_s}$ and let 
$\hat{\theta}_t = \frac{\sum_{s=1}^t \alpha_s\theta_s}{\sum_{s=1}^t \alpha_s}$.
\begin{enumerate}
\item $u_1 \in \arg\min_u h(u)$
\item $\theta_{t} \in \arg\max_{\theta \in \Theta} \langle \theta, u_t \rangle - R(\theta)$
\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \hat{\theta}_t, u \rangle$
%\item $u_{t+1} \in \arg\min_{u} h(u) + \langle \frac{1}{t} \sum_{s \leq t} \theta_s, u \rangle$
\end{enumerate}
As long as $h$ is strongly convex, we obtain the 
bound (see Proposition~\ref{cor:method-1}):
\begin{equation}
\sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T).
\end{equation}
In other words, $\hat{u}_T$ is close to being a global minimum of $L$.
However, it is often the case that $h$ is not strongly convex, whereas $R$ is strongly convex. In this case 
we may wish to use the following slightly different algorithm (called Algorithm 2):
\begin{enumerate}
\item $\theta_1 \in \arg\min_{\theta} R(\theta)$
\item $u_t \in \arg\min_{u} h(u) + \langle \theta_t, u \rangle$
%\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \frac{1}{t} \sum_{s \leq t} u_s \rangle - R(\theta)$
\item $\theta_{t+1} \in \arg\max_{\theta \in \Theta} \langle \theta, \hat{u}_t \rangle - R(\theta)$
\end{enumerate}
We then obtain the following bound 
when $R$ is strongly convex (see Proposition~\ref{cor:method-2}):
\[ \sup_{\theta \in \Theta} L(\hat{u}_T, \theta) \leq \sup_{\theta \in \Theta} L(u^*, \theta) + O(1/T). \]
Algorithm 1 and Algorithm 2 are closely related; indeed, they are dual to each other 
(performing Algorithm 1 on $L(u,\theta)$ is the same as performing Algorithm 2 on 
$-L(\theta,u)$).

@@

We now prove the convergence results cited in Section~\ref{sec:algorithm}. 
We start with two preliminary lemmas.

@@

\begin{proposition}[Convergence of Algorithm 1]
\label{prop:method-1}

@@

and $u_{t+1} \in \arg\min_u h(u) + \langle \hat{\theta}_s, u \rangle$. 
Then we have
\[ \sup_{\theta} L(\hat{u}_T, \theta) \leq \sup_{\theta} L(u^*, \theta) + \frac{\sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle}{\sum_{t=1}^T \alpha_t}. \]

@@

Lemma~\ref{lem:ftl} with $l_t(u) = \alpha_t L(u, \theta_t)$, then we get the inequality
\[ \sum_{t=1}^T \alpha_t L(u_t, \theta_t) \leq \sum_{t=1}^T \alpha_t L(u^*, \theta_t) + \sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle. \]

@@

L(\hat{u}_T, \theta) &= L\left(\frac{\sum_{t=1}^T \alpha_tu_t}{\sum_{t=1}^T \alpha_t}, \theta\right) \\
 &\leq \frac{\sum_{t=1}^T \alpha_t L(u_t, \theta)}{\sum_{t=1}^T \alpha_t} \\
 &\leq \frac{\sum_{t=1}^T \alpha_t L(u_t, \theta_t)}{\sum_{t=1}^T \alpha_t} \\
 &\leq \frac{\sum_{t=1}^T \alpha_tL(u^*, \theta_t)}{\sum_{t=1}^T \alpha_t} \\ &\phantom{=} + \frac{\sum_{t=1}^T \alpha_t (L(u_t,\theta_t)-L(u_{t+1},\theta_t))}{\sum_{t=1}^T \alpha_t} \\
 &= L(u^*, \theta_t) + \frac{\sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle)}{\sum_{t=1}^T \alpha_t} \\ &\phantom{=} + \frac{\sum_{t=1}^{T} \alpha_t(h(u_t)-h(u_{t+1}))}{\sum_{t=1}^T \alpha_t} \\
 &= L(u^*, \theta_t) + \frac{\sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle}{\sum_{t=1}^T \alpha_t} \\ &\phantom{=} + \frac{\alpha_1 h(u_1) - \alpha_T h(u_{T+1}) + \sum_{t=2}^{T} (\alpha_t-\alpha_{t-1}) h(u_t)}{\sum_{t=1}^T \alpha_t} \\
 &\leq L(u^*, \theta_t) + \frac{\sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle}{\sum_{t=1}^T \alpha_t} \\
 &\leq \sup_{\theta} L(u^*, \theta) + \frac{\sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle}{\sum_{t=1}^T \alpha_t},

@@

\label{cor:method-1}
Suppose that $h$ is $\gamma$-strongly convex and let $r = \sup_{\theta} |\langle \theta, \theta \rangle|_2$. Then 
\[ \sup_{\theta} L(\hat{u}, \theta) \leq \sup_{\theta} L(u^*, \theta) + \frac{16r^2}{\gamma T^2}. \]

@@

\|u_t-u_{t+1}\| &\leq \frac{2}{\gamma}\|\frac{\sum_{s \leq t}\alpha_s\theta_s}{\sum_{s \leq t} \alpha_s} - \frac{\sum_{s \leq t+1}\alpha_s\theta_s}{\sum_{s \leq t+1} \alpha_s}\| \\
 &= \frac{2}{\gamma}\|\frac{\alpha_{t+1}}{A_tA_{t+1}} \sum_{s \leq t} \alpha_s\theta_s - \frac{\alpha_{t+1}}{A_{t+1}} \theta_{t+1}\| \\
 &\leq \frac{2}{\gamma} \left(\frac{\alpha_{t+1}}{A_tA_{t+1}} \sum_{s \leq t} \alpha_s\|\theta_s\| + \frac{\alpha_{t+1}}{A_{t+1}} \|\theta_{t+1}\|\right) \\
 &\leq \frac{4r\alpha_{t+1}}{\gamma A_{t+1}}.

@@

\frac{1}{A_T} \sum_{t=1}^T \alpha_t \langle u_t - u_{t+1}, \theta_t \rangle &\leq \sum_{t=1}^T \frac{4r^2\alpha_{t}\alpha_{t+1}}{\gamma A_{t+1}A_T} \\
 &\leq \frac{4r^2}{\gamma A_T} \sum_{t=1}^T \frac{\alpha_{t}\alpha_{t+1}}{A_{t+1}}.

@@

In particular, if we let $\alpha_t = t$, then $A_t = \frac{t(t+1)}{2}$ and $\frac{\alpha_{t}\alpha_{t+1}}{A_{t+1}} = 2$. 
We therefore get
\begin{equation}
\frac{1}{A_T} \sum_{t=1}^T \alpha_t\langle u_t-u_{t+1}, \theta_t \rangle \leq \frac{16r^2}{T(T+1)},
\end{equation}

@@

\begin{proposition}[Convergence of Algorithm 2]
\label{prop:method-2}

@@

\label{cor:method-2}

@@

\section{Application to Herding}
\label{sec:herding}
Suppose we are given a family of features $\phi$ together 
with a known value $\bar{\phi}$ for $\bE_{\mu}[\phi]$. We 
would like to construct a distribution that approximately 
matches this distribution, i.e. one for which 
$\bE_{\hat{\mu}}[\phi] \approx \bar{\phi}$. A natural way 
to do this is by using boosted mirror descent to minimize the 
maximum mean discrepancy relative to $\phi$; 
in other words, to minimize
\[ L(\mu) = \max\limits_{i=1}^d |\bE_{\mu}[\phi_i]-\bar{\phi}_i|. \]
If we wish to write this more similarly to the original problem 
formulation, we can write
\[ L(\mu) = \sup_{\|\theta\|_1 \leq 1} \theta^T(\bE_{\mu}[\phi]-\bar{\phi}). \]
One issue is that $\|\theta\|_1$ is not strongly convex, so we relax $\|\theta\|_1$ 
to $\|\theta\|_2$ and write our loss function in the Lagrangian form:
\[ L(\mu, \theta) = \theta^T\bE_{\mu}[\phi] - \theta^T\bar{\phi} - \frac{1}{2}\|\theta\|_2^2. \]
Intriguingly, Algorithm 1 and Algorithm 2 are identical to each other for this choice of $L$ 
(up to a change of index). In this case, $\mu_t = \delta_{x_t}$ for some $x_t$, and 
the updates are the same as the standard herding updates. Let $R = \sup_{x} \|\phi(x)\|_2$; 
then using the Algorithm 2 convergence bound gives us
\begin{align*}
\sup_{\theta} L\left(\frac{1}{T} \sum_{t=1}^T \delta_{x_t}, \theta\right) &\leq \inf_{u^*} \sup_{\theta} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \sup_{\theta} \inf_{u^*} L(u^*, \theta) + \frac{2R^2\log(T+1)}{T} \\
 &= \frac{2R^2\log(T+1)}{T}.
\end{align*}
This is a slightly weaker version of the typical herding bound on MMD.

@@

Let us now consider the entropic regularizer $h(\mu) = \bE_{\mu}[\log \mu(x)]$ with $R(\theta) = \bE_{\mu}[\theta^T\bar{\phi}] + \left\{ \begin{array}{llc} \infty & : & \|\theta\|_1 > 1 \\ 0 & : & \|\theta\|_1 \leq 1 \end{array} \right.$.
If we use Method 1, we then end up with 
\[ \mu_i(x) \propto \exp(\frac{1}{i}(\sum_{j<i} \theta_j)^T\phi(x)), \]
so $h(\mu) + \bE_{\mu}[\theta^T\phi(x)] - R(\theta)$ is no longer maximized at the 
boundary. We can instead consider the following sequence of distributions:
\begin{enumerate}
\item $\theta_i \in \argmax_{\|\theta\|_1 \leq 1} \theta^T (\bE_{\mu_i}[\phi(x)]-\bar{\phi})$
\item $\mu_{i+1} \propto \exp(\frac{1}{i}(\sum_{j \leq i} \theta_j)^T\phi(x))$
\end{enumerate}
The entropic regularizer is not strongly convex but a more general analysis involving 
Bregman divergence can nevertheless yield a convergence bound in this case. We thus 
get something that looks like herding but actually gives us an exponential family 
type distribution.
